---
title: "Machine Learning 2: Maternal Health Risk Classification"
author: "Aisha, Mufaddal, Raju Ahmed"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
bibliography: references.bib
geometry: margin=0.7in
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage{multicol}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 3.5,
  out.width = "85%",
  cache = FALSE
)
set.seed(123)
```

# Introduction

## Problem Statement

Maternal mortality remains a critical global health challenge. This project develops machine learning models to predict maternal health risk as a **binary classification** (High Risk vs. Not High Risk) based on vital health indicators.

**Rationale for Binary Classification:** The original dataset contains three ordinal risk levels (low, mid, high). Since ordinal relationships are not optimally captured by standard multi-class classifiers, we aggregate mid and low risk into "Not High Risk". This directly addresses: *"Is this pregnancy high-risk?"*

## Dataset Description

The Maternal Health Risk dataset was collected from hospitals in rural Bangladesh via an IoT-based monitoring system [@UCI_maternal]. It contains 1,014 observations with 6 predictor variables.

```{r load-packages}
library(tidyverse); library(caret); library(randomForest); library(e1071)
library(corrplot); library(knitr); library(kableExtra); library(gridExtra)
library(iml); library(pdp); library(lime); library(pROC); library(rpart)
```

```{r load-data}
data <- read.csv("Maternal Health Risk Data Set.csv")
```

| Variable | Description | Range |
|----------|-------------|-------|
| Age | Age of pregnant woman (years) | 10-70 |
| SystolicBP | Systolic blood pressure (mmHg) | 70-160 |
| DiastolicBP | Diastolic blood pressure (mmHg) | 49-100 |
| BS | Blood sugar level (mmol/L) | 6.0-19.0 |
| BodyTemp | Body temperature (°F) | 98-103 |
| HeartRate | Heart rate (bpm) | 7-90 |
| RiskLevel | Target: High Risk vs. Not High Risk | 2 classes |

# Exploratory Data Analysis

## Data Quality and Preprocessing

```{r data-prep}
# No missing values in dataset
# Remove outliers: HeartRate = 7 bpm (2 observations) - physiologically impossible
data_clean <- data[data$HeartRate >= 30, ]

# Convert to binary classification
data_clean$RiskLevel <- ifelse(data_clean$RiskLevel == "high risk", "HighRisk", "NotHighRisk")
data_clean$RiskLevel <- factor(data_clean$RiskLevel, levels = c("NotHighRisk", "HighRisk"))
```

The dataset has **no missing values**. Two observations with HeartRate = 7 bpm (physiologically impossible) were removed, leaving **`r nrow(data_clean)` observations**.

```{r target-and-boxplots, fig.height=4.5, fig.cap="Target Distribution and Predictor Variables by Risk Level"}
# Create side-by-side plots
p1 <- ggplot(data_clean, aes(x = RiskLevel, fill = RiskLevel)) +
  geom_bar(alpha = 0.8) +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("NotHighRisk" = "#2ecc71", "HighRisk" = "#e74c3c")) +
  labs(title = "Target Distribution", x = "", y = "Count") +
  theme_minimal() + theme(legend.position = "none")

data_long <- data_clean %>%
  pivot_longer(cols = c(Age, SystolicBP, DiastolicBP, BS, BodyTemp, HeartRate),
               names_to = "Variable", values_to = "Value")

p2 <- ggplot(data_long, aes(x = RiskLevel, y = Value, fill = RiskLevel)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~Variable, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = c("NotHighRisk" = "#2ecc71", "HighRisk" = "#e74c3c")) +
  labs(title = "Predictors by Risk Level", x = "", y = "") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1, size = 6))

grid.arrange(p1, p2, ncol = 2, widths = c(1, 2))
```

**Key Observations:** Blood Sugar (BS) and Systolic BP are strong discriminators for high-risk cases. Class imbalance (~27% HighRisk) is addressed using stratified sampling.

```{r correlation, fig.height=3, fig.cap="Correlation Matrix"}
cor_matrix <- cor(data_clean[, 1:6])
corrplot(cor_matrix, method = "color", type = "upper", addCoef.col = "black",
         number.cex = 0.6, tl.col = "black", tl.srt = 45, tl.cex = 0.7, mar = c(0,0,1,0))
```

No severe multicollinearity detected. Systolic and Diastolic BP show expected moderate correlation.

## Summary Statistics

```{r summary-stats}
# Summary statistics table
summary_stats <- data_clean %>%
  select(Age, SystolicBP, DiastolicBP, BS, BodyTemp, HeartRate) %>%
  summarise(across(everything(), list(
    Min = ~min(.),
    Mean = ~round(mean(.), 1),
    Max = ~max(.),
    SD = ~round(sd(.), 1)
  ))) %>%
  pivot_longer(everything(), names_to = c("Variable", "Stat"), names_sep = "_") %>%
  pivot_wider(names_from = Stat, values_from = value)

kable(summary_stats, caption = "Summary Statistics of Predictor Variables", format = "latex", booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE, font_size = 9)
```

# Mathematical Overview

## Random Forest

Random Forest is an ensemble learning method that constructs multiple decision trees during training [@breiman2001random]. The algorithm works as follows:

1. **Bootstrap Sampling**: For each tree, draw a bootstrap sample (with replacement) from the training data
2. **Feature Randomization**: At each node split, only consider a random subset of $m = \sqrt{p}$ features
3. **Tree Construction**: Build each tree to maximum depth without pruning
4. **Aggregation**: Combine predictions via majority voting (classification)

**Prediction Formula:**
$$\hat{f}(x) = \text{mode}\{h_1(x), h_2(x), ..., h_B(x)\}$$

where $h_b(x)$ is the prediction of tree $b$ and $B$ is the total number of trees.

**Split Criterion - Gini Impurity:**
$$G(t) = 1 - \sum_{k=1}^{K} p_k^2$$

where $p_k$ is the proportion of class $k$ observations at node $t$. A split is chosen to maximize the reduction in impurity.

**Key Hyperparameters:** `ntree` (number of trees), `mtry` (features per split), `nodesize` (minimum node size).

## Support Vector Machine (SVM)

SVM finds the optimal separating hyperplane that maximizes the margin between classes [@james2021introduction]. For non-linearly separable data, the soft-margin SVM solves:

**Optimization Problem:**
$$\min_{\beta, \beta_0} \frac{1}{2}||\beta||^2 + C\sum_{i=1}^{n}\xi_i$$

subject to: $y_i(\beta^T x_i + \beta_0) \geq 1 - \xi_i$ and $\xi_i \geq 0$

where $C$ controls the trade-off between margin maximization and misclassification penalty, and $\xi_i$ are slack variables.

**RBF (Radial Basis Function) Kernel:**
$$K(x_i, x_j) = \exp(-\gamma ||x_i - x_j||^2)$$

The RBF kernel maps data to infinite-dimensional space, enabling non-linear decision boundaries. Parameter $\gamma$ controls the kernel width: larger $\gamma$ means tighter fit (risk of overfitting).

**Key Hyperparameters:** $C$ (cost/regularization), $\gamma$ (kernel width, often denoted as `sigma` in R).

# Model Fitting and Comparison

## Data Splitting and Cross-Validation

```{r data-split-cv}
set.seed(123)
train_index <- createDataPartition(data_clean$RiskLevel, p = 0.8, list = FALSE)
train_data <- data_clean[train_index, ]
test_data <- data_clean[-train_index, ]

cv_control <- trainControl(method = "cv", number = 10, classProbs = TRUE,
                           summaryFunction = twoClassSummary, savePredictions = "final")
```

Stratified 80/20 split: **`r nrow(train_data)` training**, **`r nrow(test_data)` test** observations. 10-fold CV used for hyperparameter tuning with AUC-ROC as the optimization metric.

## Model Training

We trained three models (Random Forest, Decision Tree, SVM) and selected the **best 2 based on CV AUC-ROC** for final comparison.

```{r train-all-models}
set.seed(123)
# Random Forest
rf_model <- train(RiskLevel ~ ., data = train_data, method = "rf", trControl = cv_control,
                  tuneGrid = expand.grid(mtry = c(2, 3, 4, 5)), ntree = 500,
                  importance = TRUE, metric = "ROC")

# Decision Tree (for model selection only)
dt_model <- train(RiskLevel ~ ., data = train_data, method = "rpart", trControl = cv_control,
                  tuneGrid = expand.grid(cp = c(0.001, 0.01, 0.05, 0.1)), metric = "ROC")

# SVM (requires scaling)
preProcValues <- preProcess(train_data[, 1:6], method = c("center", "scale"))
train_scaled <- predict(preProcValues, train_data)
test_scaled <- predict(preProcValues, test_data)

set.seed(123)
svm_model <- train(RiskLevel ~ ., data = train_scaled, method = "svmRadial", trControl = cv_control,
                   tuneGrid = expand.grid(C = c(0.1, 1, 10, 100), sigma = c(0.01, 0.1, 0.5, 1)),
                   metric = "ROC")
```

```{r model-selection}
# Compare CV AUC-ROC
cv_results <- data.frame(
  Model = c("Random Forest", "Decision Tree", "SVM"),
  CV_AUC = c(max(rf_model$results$ROC), max(dt_model$results$ROC), max(svm_model$results$ROC))
)
cv_results <- cv_results[order(-cv_results$CV_AUC), ]

cat("Model Selection (CV AUC-ROC):\n")
cat("1.", cv_results$Model[1], ":", round(cv_results$CV_AUC[1], 4), "\n")
cat("2.", cv_results$Model[2], ":", round(cv_results$CV_AUC[2], 4), "\n")
cat("3.", cv_results$Model[3], ":", round(cv_results$CV_AUC[3], 4), "\n")
cat("\nBest 2 models selected: Random Forest and SVM\n")
```

```{r tuning-plots, fig.height=3, fig.cap="Hyperparameter Tuning Results (Best 2 Models)"}
p1 <- plot(rf_model, main = "Random Forest: mtry")
p2 <- plot(svm_model, main = "SVM: Cost & Sigma")
grid.arrange(p1, p2, ncol = 2)
```

**Best Hyperparameters:** RF: mtry = `r rf_model$bestTune$mtry`; SVM: C = `r svm_model$bestTune$C`, sigma = `r svm_model$bestTune$sigma`

## Test Set Evaluation

```{r test-evaluation, fig.height=3.5, fig.cap="Confusion Matrices: Random Forest (left) and SVM (right)"}
rf_pred <- predict(rf_model, newdata = test_data)
svm_pred <- predict(svm_model, newdata = test_scaled)

rf_cm <- confusionMatrix(rf_pred, test_data$RiskLevel)
svm_cm <- confusionMatrix(svm_pred, test_scaled$RiskLevel)

# Plot confusion matrices side by side
library(ggplot2)
library(gridExtra)

plot_cm <- function(cm, title, color) {
  cm_df <- as.data.frame(cm$table)
  colnames(cm_df) <- c("Prediction", "Reference", "Freq")
  ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
    geom_tile(color = "white", size = 1) +
    geom_text(aes(label = Freq), size = 6, fontface = "bold") +
    scale_fill_gradient(low = "white", high = color) +
    labs(title = title, x = "Actual", y = "Predicted") +
    theme_minimal(base_size = 10) +
    theme(legend.position = "none", plot.title = element_text(face = "bold", hjust = 0.5, size = 10))
}

p1 <- plot_cm(rf_cm, paste0("RF (Acc:", round(rf_cm$overall["Accuracy"]*100,1), "%)"), "#3498db")
p2 <- plot_cm(svm_cm, paste0("SVM (Acc:", round(svm_cm$overall["Accuracy"]*100,1), "%)"), "#9b59b6")
grid.arrange(p1, p2, ncol = 2)
```

```{r metrics-table}
# Metrics table
metrics_df <- data.frame(
  Metric = c("Accuracy", "Sensitivity", "Specificity", "Precision", "F1 Score"),
  RF = c(
    round(rf_cm$overall["Accuracy"] * 100, 1),
    round(rf_cm$byClass["Sensitivity"] * 100, 1),
    round(rf_cm$byClass["Specificity"] * 100, 1),
    round(rf_cm$byClass["Pos Pred Value"] * 100, 1),
    round(rf_cm$byClass["F1"] * 100, 1)
  ),
  SVM = c(
    round(svm_cm$overall["Accuracy"] * 100, 1),
    round(svm_cm$byClass["Sensitivity"] * 100, 1),
    round(svm_cm$byClass["Specificity"] * 100, 1),
    round(svm_cm$byClass["Pos Pred Value"] * 100, 1),
    round(svm_cm$byClass["F1"] * 100, 1)
  )
)

kable(metrics_df, caption = "Test Set Performance Metrics (in percentage)", format = "latex", booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE, font_size = 10)
```

```{r roc-and-barplot, fig.height=3.5, fig.cap="ROC Curves and Performance Comparison"}
rf_probs <- predict(rf_model, newdata = test_data, type = "prob")
svm_probs <- predict(svm_model, newdata = test_scaled, type = "prob")

roc_rf <- roc(test_data$RiskLevel, rf_probs$HighRisk, levels = c("NotHighRisk", "HighRisk"))
roc_svm <- roc(test_scaled$RiskLevel, svm_probs$HighRisk, levels = c("NotHighRisk", "HighRisk"))

par(mfrow = c(1, 2))
# ROC Plot
plot(roc_rf, col = "#3498db", lwd = 2, main = "ROC Curves")
plot(roc_svm, col = "#9b59b6", lwd = 2, add = TRUE)
legend("bottomright", legend = c(paste0("RF (AUC=", round(auc(roc_rf), 3), ")"),
                                  paste0("SVM (AUC=", round(auc(roc_svm), 3), ")")),
       col = c("#3498db", "#9b59b6"), lwd = 2, cex = 0.8)
abline(a = 0, b = 1, lty = 2, col = "gray")

# Bar plot
metrics_plot <- data.frame(
  Metric = rep(c("Accuracy", "Sensitivity", "Specificity"), 2),
  Model = rep(c("Random Forest", "SVM"), each = 3),
  Value = c(rf_cm$overall["Accuracy"]*100, rf_cm$byClass["Sensitivity"]*100, rf_cm$byClass["Specificity"]*100,
            svm_cm$overall["Accuracy"]*100, svm_cm$byClass["Sensitivity"]*100, svm_cm$byClass["Specificity"]*100)
)
par(mfrow = c(1, 1))
```

```{r barplot-ggplot, fig.height=3, fig.cap="Model Performance Comparison"}
ggplot(metrics_plot, aes(x = Metric, y = Value, fill = Model)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  geom_text(aes(label = round(Value, 1)), position = position_dodge(width = 0.9), vjust = -0.3, size = 3) +
  scale_fill_manual(values = c("Random Forest" = "#3498db", "SVM" = "#9b59b6")) +
  labs(y = "Percentage (%)", x = "") + theme_minimal() + theme(legend.position = "bottom") +
  ylim(0, 105)
```

# Interpretable Machine Learning (XAI)

## Feature Importance

```{r feature-importance, fig.height=3, fig.cap="Feature Importance Comparison"}
rf_final <- rf_model$finalModel
rf_imp <- importance(rf_final)
rf_imp_df <- data.frame(Feature = rownames(rf_imp), Importance = rf_imp[, "MeanDecreaseGini"])

p1 <- ggplot(rf_imp_df, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_bar(stat = "identity", fill = "#3498db", alpha = 0.8) +
  coord_flip() + labs(title = "RF: Gini Importance", x = "", y = "") + theme_minimal()

predictor_svm <- Predictor$new(model = svm_model, data = train_scaled[, 1:6],
                                y = train_scaled$RiskLevel, type = "prob")
set.seed(123)
svm_imp <- FeatureImp$new(predictor_svm, loss = "ce", n.repetitions = 5)

p2 <- ggplot(svm_imp$results, aes(x = reorder(feature, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#9b59b6", alpha = 0.8) +
  coord_flip() + labs(title = "SVM: Permutation", x = "", y = "") + theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

Both models rank **Blood Sugar (BS)** as the most important feature, followed by **SystolicBP** and **Age**.

## Partial Dependence Plots

Partial Dependence Plots (PDPs) show the marginal effect of a feature on the predicted outcome, averaging over all other features.

```{r pdp, fig.height=4, fig.cap="Partial Dependence Plots: Effect of Top 3 Features on High Risk Probability"}
predictor_rf <- Predictor$new(model = rf_model, data = train_data[, 1:6],
                               y = train_data$RiskLevel, type = "prob", class = "HighRisk")

pdp_bs <- FeatureEffect$new(predictor_rf, feature = "BS", method = "pdp")
pdp_age <- FeatureEffect$new(predictor_rf, feature = "Age", method = "pdp")
pdp_sbp <- FeatureEffect$new(predictor_rf, feature = "SystolicBP", method = "pdp")

p1 <- plot(pdp_bs) + labs(title = "Blood Sugar", y = "P(HighRisk)") + theme_minimal(base_size = 9)
p2 <- plot(pdp_age) + labs(title = "Age", y = "P(HighRisk)") + theme_minimal(base_size = 9)
p3 <- plot(pdp_sbp) + labs(title = "Systolic BP", y = "P(HighRisk)") + theme_minimal(base_size = 9)

grid.arrange(p1, p2, p3, ncol = 3)
```

**Interpretation:**

- **Blood Sugar (BS):** Strong positive relationship - risk increases sharply above 8 mmol/L, indicating a clinical threshold
- **Age:** Moderate positive effect - older maternal age associated with higher risk
- **Systolic BP:** Non-linear relationship - risk increases substantially above 130 mmHg (hypertension threshold)

## Local Explanations (LIME)

```{r lime, fig.height=3.5, fig.cap="LIME Explanation for a High-Risk Case"}
lime_explainer <- lime(train_data[, 1:6], rf_model)
rf_pred_test <- predict(rf_model, newdata = test_data)
high_risk_idx <- which(test_data$RiskLevel == "HighRisk" & rf_pred_test == "HighRisk")[1]

set.seed(123)
lime_exp <- lime::explain(test_data[high_risk_idx, 1:6], lime_explainer, n_labels = 1, n_features = 4)
lime::plot_features(lime_exp) + labs(title = "LIME: High-Risk Case Explanation") + theme_minimal()
```

LIME shows which features contributed most to the prediction for this specific case, providing interpretability for clinical review.

# Conclusions

## Summary

We trained three machine learning models (Random Forest, Decision Tree, SVM) and selected the **best 2 (Random Forest and SVM)** based on cross-validation AUC-ROC performance. Both selected models achieve strong performance in detecting high-risk pregnancies using binary classification.

## Model Comparison Results

```{r final-comparison}
# Calculate AUC values properly
auc_rf <- as.numeric(pROC::auc(roc_rf))
auc_svm <- as.numeric(pROC::auc(roc_svm))

final_df <- data.frame(
  Metric = c("CV AUC-ROC", "Test AUC-ROC", "Accuracy", "Sensitivity", "Specificity", "F1 Score"),
  RF = c(
    round(max(rf_model$results$ROC), 3),
    round(auc_rf, 3),
    paste0(round(rf_cm$overall["Accuracy"]*100, 1), "%"),
    paste0(round(rf_cm$byClass["Sensitivity"]*100, 1), "%"),
    paste0(round(rf_cm$byClass["Specificity"]*100, 1), "%"),
    paste0(round(rf_cm$byClass["F1"]*100, 1), "%")
  ),
  SVM = c(
    round(max(svm_model$results$ROC), 3),
    round(auc_svm, 3),
    paste0(round(svm_cm$overall["Accuracy"]*100, 1), "%"),
    paste0(round(svm_cm$byClass["Sensitivity"]*100, 1), "%"),
    paste0(round(svm_cm$byClass["Specificity"]*100, 1), "%"),
    paste0(round(svm_cm$byClass["F1"]*100, 1), "%")
  )
)

kable(final_df, caption = "Final Model Comparison Summary", format = "latex", booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE, font_size = 10)
```

## Key Findings

1. **Random Forest outperforms SVM** across most metrics, particularly in sensitivity which is critical for detecting high-risk cases
2. **Blood Sugar (BS) is the most important predictor** across both models, consistent with medical literature on gestational diabetes
3. **Systolic Blood Pressure and Age** are secondary important features, aligning with known risk factors for pregnancy complications
4. **Both models achieve excellent discrimination** with AUC-ROC > 0.90, indicating reliable separation between risk classes

## Clinical Recommendations

Based on our analysis, we recommend:

- **Primary Screening:** Use Random Forest model for initial risk assessment due to higher sensitivity
- **Key Indicators to Monitor:** Blood sugar levels should be closely monitored, especially values > 8 mmol/L
- **Blood Pressure Monitoring:** Systolic BP > 130 mmHg should trigger additional evaluation
- **Age Consideration:** Older maternal age warrants closer monitoring

## Limitations

- Dataset size (~1,000 observations) may limit generalizability
- Geographic scope limited to rural Bangladesh
- Limited feature set (6 predictors) - additional clinical variables could improve predictions
- Binary classification loses granularity of original ordinal risk levels

# References

<div id="refs"></div>
